Exploration of Red Wine Quality by Cady Patino
========================================================


> **Tip**: One of the requirements of this project is that your code follows
good formatting techniques, including limiting your lines to 80 characters or
less. If you're using RStudio, go into Preferences \> Code \> Display to set up
a margin line to help you keep track of this guideline!

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(gridExtra)
library(corrplot)
library(GGally)
library(psych)
library(dplyr)
library(RColorBrewer)
library(MASS)
library(memisc)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
wine <- read.csv('WineQualityReds.csv')
```

> This report explores a dataset containing 11 variables of the chemical properties of red wines from the Portuguese Vinho Verde region and their corresponding quality rating. The quality rating for each wine is the median grade given by at least three wine experts and has a value between 0 (very bad) and 10 (very excellent).

# Univariate Plots Section

```{r echo=FALSE, Univariate_Plots}
str(wine)
```

```{r}
wine <- wine[ , names(wine) != 'X']
wine$quality.factor <- factor(wine$quality)
summary(wine)
```

> This data set contains 1599 observations of 13 variables. The first variable, X, appears to be the index column, so I removed it from the dataset before performing the summary. The next 11 variables are numbers pertaining to the chemical properties of wine, and the final variable is the quality rating. I converted the quality rating from an integer to a factor and added it as a new variable, quality.factor. This way I can treat the quality rating as both continous and as a factor in further analyses. 

```{r}
qplot(x=quality.factor, data=wine)
```

> From the above plot, it is clear that all quality values in this data set are integers. The majority of quality ratings in this data set are 5 or 6, and 3 and 8 are the least common ratings.


```{r}
qplot(x=fixed.acidity, data=wine, binwidth=0.1) +
  scale_x_continuous(breaks=seq(4,16,1))
```

```{r}
qplot(x=volatile.acidity, data=wine, binwidth=0.01) +
  scale_x_continuous(breaks=seq(0,1.6,0.1))
```

```{r}
qplot(x=citric.acid, data=wine, binwidth=0.01) +
  scale_x_continuous(breaks=seq(0,1,0.1))
```

```{r}
qplot(x=pH, data=wine, binwidth=0.01) +
  scale_x_continuous(breaks=seq(0,4.01,0.1))
```

> Above are the histograms related to acidity and pH. The bulk of the fixed acidity data is between 7 and 9. The bulk of the volatile acidity is between 0.3 and 0.7. These two variables follow a fairly normal distribution with some outliers on the high end. The data for citric acid, on the other hand, is much more spread out, with peaks at 0 and about 0.49. The majority of citric acid data falls between 0 and 0.5. Lastly, the plot for pH appears normally distributed with just a few outliers around 2.75 and 3.8-4.0. The bulk of the pH data falls between 3.1 and 3.5 with a peak around 3.3.

```{r}
qplot(x=residual.sugar, data=wine, binwidth=0.1) +
  scale_x_continuous(breaks=seq(0,16,1))
```

```{r}
qplot(x=residual.sugar, data=wine, binwidth=0.01) +
  scale_x_log10()

```

> Above, I plotted histograms of residual sugar and the log transformation of residual sugar due to its long-tailed nature. The peak of the data is around 2-2.5. The transformed data still has a tail due to far outliers out to 16.

```{r}
qplot(x=chlorides, data=wine, binwidth=0.005) +
  scale_x_continuous(breaks=seq(0,0.62,0.05))
```

```{r}
qplot(x=chlorides, data=wine, binwidth=0.02) +
  scale_x_log10()
```

> Similarly to the sugar data, I plotted histograms of the chloride data and its log transofmration due to its right-skewdness and long tail. The bulk of the data is between 0.06 and 0.1, with a peak of about 0.8.

```{r}
qplot(x=free.sulfur.dioxide, data=wine, binwidth=1) +
  scale_x_continuous(breaks=seq(0,72,5))
```

```{r}
qplot(x=free.sulfur.dioxide, data=wine, binwidth=0.02) +
  scale_x_log10(breaks=seq(0,80, 10))
```


```{r}
qplot(x=total.sulfur.dioxide, data=wine, binwidth=2) +
  scale_x_continuous(breaks=seq(0,290,20))
```

```{r}
qplot(x=total.sulfur.dioxide, data=wine, binwidth=0.02) +
  scale_x_log10(breaks=seq(0,300,25))
```

```{r}
qplot(x=sulphates, data=wine, binwidth=0.02) +
  scale_x_continuous(breaks=seq(0,2,0.2))
```

```{r}
qplot(x=sulphates, data=wine, binwidth= 0.01) +
  scale_x_log10(breaks=seq(0,2,0.2))
```

> Above, I've plotted histograms of the data pertaining to sulfer dioxide and sulphates. I plotted the log transformation of each of these variables as well, since the data are right-skewed and long-tailed. For free sulfur dioxide and total sulfur dioxide, after the transformation you can see some discreteness in the left side of the graphs, and the data are fairly wide spread. For free sulfur dioxide, the bulk of the data is between 5 and 20, with a peak around 6. For total sulfur dioxide, the bulk of the data is between 15 and 75, with a peak around 30. Lastly, for sulphates (a wine additive that contributes to sulfur dioxide gas levels), the data is more narrowly distributed, with the majority of the data falling between 0.5 and 0.8. The peak for sulphates appears to be around 5.8. 

```{r}
qplot(x=density, data=wine, binwidth=0.0002) +
  scale_x_continuous(breaks=seq(0.990,1.004, 0.001))
```

> Above, the density histogram is already normally distributed, with the majority of the data falling between 0.995 and 0.999. Practically, this means that there is very little difference in density across the different wines evaluated for this dataset. I wonder if these very small differences affect quality ratings. My guess is that our mouths can't tell the difference between this very small range of densities.

```{r}
qplot(x=alcohol, data=wine, binwidth= 0.1) +
  scale_x_continuous(breaks=seq(8,15,0.5))
```

> Lastly, I plotted the histogram for alcohol content. This variable has a fairly large spread. The bulk of the data is between 9.25 and 11.5, and some high outliers can be seen around 14 and about 14.8. Some low outliers can be seen as well, around 8.5. I would assume that alcohol content probably has a large affect on wine quality since alcohol content can greatly affect the taste.

> I'm curious as to how each of these variables affects the quality of wine as perceived by experts. I want to now examine how these variables look for the highest quality wines (rating of 7 or 8) compared to the lowest quality wines (rating of 3 or 4) in this dataset. Note, the high quality subset has 217 samples, while the low quality subset only has 63 samples.

```{r}
high_quality <- subset(wine, quality == 7 | quality == 8)
low_quality <- subset(wine, quality == 3 | quality == 4)

p1 <- qplot(x=fixed.acidity, data=high_quality)
p2 <- qplot(x=fixed.acidity, data=low_quality)
p3 <- qplot(x=volatile.acidity, data=high_quality)
p4 <- qplot(x=volatile.acidity, data=low_quality)
p5 <- qplot(x=citric.acid, data = high_quality)
p6 <- qplot(x=citric.acid, data=low_quality)
p7 <- qplot(x=pH, data=high_quality)
p8 <- qplot(x=pH, data=low_quality)

grid.arrange(p1, p2, p3, p4, p5, p6, p7, p8, ncol=2)
```

> Above, I plotted the subsets of high quality (left plots) and low quality wines (right plots). Let's look at the summaries as well.

```{r}
summary(high_quality$fixed.acidity)
summary(low_quality$fixed.acidity)
```

> Above are the summaries for fixed acidity for the high quality (first) and low quality (second) wines. The highest quality wines have a greater mean and median fixed acidity than the lower quality wines, as well as a higher max value.

```{r}
summary(high_quality$volatile.acidity)
summary(low_quality$volatile.acidity)
```

> Above are the summaries for volatile acidity for the high quality (first) and low quality (second) wines. The lowest quality wines have a much higher mean and median volatile acidity than the highest quality wines.

```{r}
summary(high_quality$citric.acid)
summary(low_quality$citric.acid)
```

> Above are the summaries for citric acid content in the wine samples from the high quality (first) and low quality (second) data subsets. The high quality subset has a greater mean and median citric acid content than the low quality subset.

```{r}
summary(high_quality$pH)
summary(low_quality$pH)
```

> Interestingly, the pH summaries for the high quality (first) and low quality (second) subsets are very similar. They have similar mean and median values, with the low quality wines having just slightly greater values.

```{r}
p1 <- qplot(x=residual.sugar, data=high_quality) + scale_x_log10()
p2 <- qplot(x=residual.sugar, data=low_quality) + scale_x_log10()
p3 <- qplot(x=chlorides, data=high_quality) + scale_x_log10()
p4 <- qplot(x=chlorides, data=low_quality) + scale_x_log10()

grid.arrange(p1, p2, p3, p4, ncol = 2)
```

> Above are the histograms for sugar and salts in the wine samples, broken up into subsets of high quality wines (left) and low quality wines (right).

```{r}
summary(high_quality$residual.sugar)
summary(low_quality$residual.sugar)
```

> The summary above is for residual sugar for the high quality (first) and low quality (second) data subsets. The mean and median residual sugar content are similar for both subsets; they are just slightly higher for the high quality wines. The low quality subset has a much higher max value, however.

```{r}
summary(high_quality$chlorides)
summary(low_quality$chlorides)
```

> Above, looking at the summary of chloride content, we can see that the lower quality wines (second) have a slightly higher median and mean than the higher quality wines (first).

```{r}
p1 <- qplot(x= free.sulfur.dioxide, data = high_quality) + scale_x_log10()
p2 <- qplot(x= free.sulfur.dioxide, data = low_quality) + scale_x_log10()
p3 <- qplot(x= total.sulfur.dioxide, data = high_quality) + scale_x_log10()
p4 <- qplot(x= total.sulfur.dioxide, data = low_quality) + scale_x_log10()
p5 <- qplot(x= sulphates, data = high_quality) + scale_x_log10()
p6 <- qplot(x= sulphates, data = low_quality) + scale_x_log10()

grid.arrange(p1,p2,p3,p4,p5,p6, ncol = 2)
```

> Above are the histograms for the variables relating to sulfur dioxide and sulphates, broken up into subsets of high quality wines (left) and low quality wines (right).

```{r}
summary(high_quality$free.sulfur.dioxide)
summary(low_quality$free.sulfur.dioxide)
```

> From the summary above, we can see that the mean and median of free sulfur dioxide is higher in the high quality wines (first) than the low quality wines (second).

```{r}
summary(high_quality$total.sulfur.dioxide)
summary(low_quality$total.sulfur.dioxide)
```

> The summary above shows that the mean of total sulfur dioxide content is nearly the same for the high (first) and low (second) quality wines, but the median is slightly higher for the high quality wines.

```{r}
summary(high_quality$sulphates)
summary(low_quality$sulphates)
```

> And lastly in regards to sulfur dioxide and sulphates, the above summary shows that the high quality wines (first) have a higher mean and median sulphate concentration than the low quality wines (second).

```{r}
p1 <- qplot(x= density, data = high_quality)
p2 <- qplot(x= density, data = low_quality)
p3 <- qplot(x= alcohol, data = high_quality)
p4 <- qplot(x= alcohol, data = low_quality)

grid.arrange(p1, p2, p3, p4, ncol=2)
```

> Above is our last set of histogram plots of the high quality wines (left) compared to the low quality wines (right), this time for density and alcohol content. 

```{r}
summary(high_quality$density)
summary(low_quality$density)
```

> The mean and median densities of the hiqh quality (first) and low quality (second) wines are practically the same, with the low quality wines having only a slightly greater density.

```{r}
summary(high_quality$alcohol)
summary(low_quality$alcohol)
```

> The above summary shows that the mean and median of alcohol content are greater for the high quality wines (first) than the low quality wines (second).

# Univariate Analysis

### What is the structure of your dataset?
There are 1599 observations in this dataset with 12 features of wine: fixed acidity, volatile acidity, citric acid, residual sugar, chlorides, free sulfur dioxide, total sulfur dioxide, density, pH, sulphates, alcohol, and quality. The quality variable is of type integer and the rest are of type numeric.

Other observations:
- The median quality rating is 6, and the range is from 3 to 8. Most wines have a quality rating of 5 or 6, and 25% of wines have a quality rating above 6.
- Most wines have a fixed acidity between 7 and 9, and a volatile acidity of 0.3-0.7.
- Citric acid concentration tends to be wide spread, with the majority between 0 and 0.5 and a median of of 0.26.
- The mean and median pH is 3.31. Most wines have a pH between 3.1 and 3.5.
- Most wines have a residual sugar content between 2 and 2.5 and a chloride concentration between 0.06 and 0.1.
- Free sulfur dioxide and total sulfur dioxide are also widespread, with the majority of wines falling in the range of 5-20, and 15-75 respectively. Median free sulfur dioxide is 14 and median total sulfur dioxide is 38.
- Most wines have a sulphate concentration between 0.5 and 0.8.
- The median alcohol content is 10.2, with most wines having alcohol content between 9.25 and 11.5. 
- Density has a very narrow range (~0.01) with a median of 0.9968.

### What is/are the main feature(s) of interest in your dataset?
The main feature of interest in this dataset is quality. I'm interested in seeing how the chemical properties of wine affect the quality rating of wine given by experts. 

### What other features in the dataset do you think will help support your \ investigation into your feature(s) of interest?
I think the features of fixed acidity, volatile acidity, citric acid, chlorides, free sulfur dioxide, sulphates, and alcohol are most likely to affect the quality of wine, based on my initial comparisons of high quality wines (7 or 8) vs. low quality wines (3 or 4). My guess is that volatile acidity, citric acid content, sulfur dioxide content, and alcohol will have the greatest affect on quality because these properties tend to directly alter taste and smell.  

### Did you create any new variables from existing variables in the dataset?
I did not create any new variables at this point.

### Of the features you investigated, were there any unusual distributions? \ Did you perform any operations on the data to tidy, adjust, or change the form \ of the data? If so, why did you do this?

The distributions for residual sugar, chlorides, and sulphates were originally long-tailed on the right, and the distributions for free sulfur dioxide and total sulfur dioxide were skewed to the right. I decided to log-transform these distributions so that they would more closely resemble that of a normal distribution.

The data loaded in was already tidy and in the form that I wanted. After plotting initial histograms, I created two subsets of the data by selecting only the highest quality wines (quality rating of 7 or 8)  and the lowest quality wines (quality rating of 3 or 4) from the original dataset and comparing their distributions and summaries for each feature.

# Bivariate Plots Section

```{r echo=FALSE, Bivariate_Plots}
wine.corr <- cor(wine[1:12])
corrplot(wine.corr, method='circle')
```

> Above is a correlation matrix of the continuous variables of the dataset. Some strong correlations that were expected are fixed acidity vs. citric acid, fixed acidity vs. pH, and citric acid vs. pH. In addition, free sulfur dioxide and total sulfur dioxide have a strong correlation which makes sense because total sulfur dioxide is the total of free and bound sulfur dioxide. Surpringly, sulphate concentration is not correlated with free or total sulfur dioxide in this dataset; I expected these to be correlated because sulphates are a wine additive that can contribute to sulfur dioxide gas levels. Density appears to be moderately correlated with alcohol and fixed acidity. Lastly, citric acid has a negative correlation with volatile acidity, which I did not expect.

```{r}
pairs.panels(wine)
```

> Above is a scatterplot matrix of all the variables in the data set, providing an additional visualization of the correlations. Most importantly, quality appears to be most correlated with volatile acidity and alcohol, followed by sulphates and citric acid. I now want to take a closer look at these correlations as well as the correlation between some of the supporting variables mentioned above.

```{r}
ggplot(aes(x=quality.factor, y=volatile.acidity), data=wine) + geom_boxplot()
```

```{r}
tapply(wine$volatile.acidity, wine$quality.factor, summary)
```

> Quality of wine tends to increase with decreasing volatile acidity values. The lowest quality ratings (3 and 4) have the highest median volatile acidity values and the largest interquartile ranges.

```{r}
ggplot(aes(x=quality.factor, y=alcohol), data = wine) +
  geom_boxplot()
```

```{r}
tapply(wine$alcohol, wine$quality.factor, summary)
```

> Higher quality wines tend to have higher alcohol contents. The group with quality 5 is the exception to this trend, and has the highest max value and the most outliers. 

```{r}
ggplot(aes(x=quality.factor, y = citric.acid), data = wine) +
  geom_boxplot()
```

```{r}
tapply(wine$citric.acid, wine$quality.factor, summary)
```

> Median citric acid concentration increases as quality increases. Citric acid can add freshness and flavor to wines, so it makes sense that using it as an additive can increase the perceived quality of the wine.

```{r}
ggplot(aes(x=quality.factor, y = sulphates), data = wine) +
  geom_boxplot()
```
```{r}
ggplot(aes(x=quality.factor, y = sulphates), data = subset(wine, sulphates < 1.5)) +
  coord_cartesian(ylim=c(0.3,1)) +
  geom_boxplot()
```

```{r}
tapply(wine$sulphates, wine$quality.factor, summary)
```

> The quality rating of wine tends to increase with increasing sulphate levels. The median sulphate levels are higher in the highest rated wines than in the lowest rated wines. Potassium sulphate is a wine additive that can contribute to sulfur dioxide gas levels, which helps to prevent oxidation and to maintain a wine's freshness. Since sulphates seem to have an effect on quality, let's also look at the relationship between total sulfur dioxide and quality.

```{r}
p1 <- ggplot(aes(x=quality.factor, y = total.sulfur.dioxide), data = wine) +
  coord_cartesian(ylim=c(0,100)) +
  geom_boxplot()

p2 <- ggplot(aes(x=quality.factor, y = free.sulfur.dioxide), data = wine) +
  coord_cartesian(ylim=c(0,100)) +
  geom_boxplot()

grid.arrange(p1,p2,ncol=2)
```

> Interestingly, free and total sulfur dioxide levels do not appear to affect the quality of wine, despite the fact that sulphates do. This goes against my intuition since sulphates contribute to sulfur dioxide levels, and sulfur dioxide levels can become evident in the taste of wine. However, the dataset information states that free sulfur dioxide levels below 50ppm are generally undetected by the nose and taste. The majority of wines at all quality levels have a free sulfur dioxide level below this threshold. 

```{r}
p1 <- ggplot(aes(x=sulphates, y=total.sulfur.dioxide), data=wine) +
  geom_point(alpha=1/5)

p2 <- ggplot(aes(x=sulphates, y=free.sulfur.dioxide), data=wine) +
  geom_point(alpha=1/5)

grid.arrange(p1,p2,ncol=2)
```

> Again, it comes as a bit of a surprise that neither free or total sulfur dioxide is correlated to sulphates as shown above. After doing some more research however, I found that the alcohol fermentation process produces sulfur dioxide in amounts ranging from 10 to over 100 ppm. Therefore, even in wines without a sulphate additive, sulfur dioxide exists.

```{r}
ggplot(aes(x=citric.acid, y=fixed.acidity), data=wine) + 
  geom_point(alpha=1/4) +
  geom_smooth(method='lm')
```

> Citric acid has a positive correlation with fixed.acidity that appears to be linear. This is logical since citric acid would contribute to the acidity of the wine.

```{r}
ggplot(aes(x=citric.acid, y=volatile.acidity), data=wine) + 
  geom_point(alpha=1/4)
```

> Citric acid also appears to have a correlation with volatile acidity. However, unlike with fixed acidity, this is an inverse relationship, and is not quite linear. According to the dataset information page, volatile acidity is the measure of acetic acid which evaporates readily. I wonder why citric acid would affect this.

```{r}
ggplot(aes(x=fixed.acidity, y=pH), data=wine) + geom_point(alpha=1/4)
```

```{r}
ggplot(aes(x=log(fixed.acidity), y=pH), data=wine) + 
  geom_point(alpha=1/4) +
  geom_smooth(method='lm')
```

> As intution would suggest, we can see that pH decreases as fixed acidity increases. When plotting pH against the log of fixed acidity, this relationship appears linear. 

```{r}
ggplot(aes(x=citric.acid, y=pH), data=wine) + 
  geom_point(alpha=1/4) + 
  geom_smooth(method=lm)
```

> A similar relationship can be seen between pH and citric acid above.

```{r}
ggplot(aes(x=alcohol, y=density), data=wine) +
  geom_jitter(alpha=1/4)
```

> As alcohol content increases, density appears to decrease. However, these density ranges are miniscule and are most likely undetectable, which is why we can conclude it is the alcohol itself affecting quality and not the slight differences in density.

```{r}
ggplot(aes(x=fixed.acidity, y=density), data=wine) + 
  geom_point(alpha=1/4)
```

> Fixed acidity has the opposite effect on density as alcohol does. As fixed acidity increases, density increases slightly.

# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \ investigation. How did the feature(s) of interest vary with other features in \ the dataset?
I found that quality increases as volatile acidity (acetic acid) decreases. With a correlation coefficient of 0.39, this can be considered a weak to moderate correlation. This effect is intuitive since high amounts of acetic acid in wine can lead to an unpleasant vinegar taste. 

In addition, higher quality wines are associated with higher alcohol contents. Alcohol content explains the largest percentage of variance in wine quality, with a correlation coefficient of 0.48.

Citric acid levels and sulphate levels also appeared to have a positive effect on quality, although these correlations were weaker (0.23 and 0.25 respectively). Sulfur dioxide levels, however, do not appear to affect quality, at least in the range measured in this dataset, despite the fact that sulfur dioxide gas is a byproduct of sulphates. 

### Did you observe any interesting relationships between the other features \ (not the main feature(s) of interest)?
One interesting relationship that I did not expect was the inverse relationship between citric acid and volatile acidity (acetic acid concentration). From doing some research, I found that during the malolatic fermentation process (which is often performed after the primary fermentation process), the metabolism of citric acid occurs; the citric-sugar co-metabolism can increase the formation of volatile acid in wine. From this information I would assume a positive relationship would exist between citric acid and volatile acidity.

Another interesting, unexpected relationship is the positive relationship between fixed acidity and density. I am not sure why this occurs.

Lastly, I expected that sulphates would have a positive correlation to free or total suflur dioxide levels since sulfur dioxide is a byproduct. However, there was no apparent relationship between these variables, suggesting that the sulfur dioxide levels produced by fermentation probably overshadows those produced by sulphates in many cases.

http://wineserver.ucdavis.edu/industry/enology/methods_and_techniques/reagents/citric_acid.html

### What was the strongest relationship you found?
The strongest relationship I found was that of fixed acidity and pH, with a correlation coefficient of 0.68 (strong). The strongest relationship I found regarding the main feature of interest, quality, was quality vs. alcohol content with a correlation coefficient of 0.48 (moderate).

# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

> Based on what I found above, I first want to look at the effects of volatile acidity and alcohol together with quality, as well as some density plots.

```{r echo=FALSE, Multivariate_Plots}
#create buckets for alcohol content
wine$alcohol.bucket <- cut(wine$alcohol, breaks=c(8,9.25,10,11,12.25,15))
table(wine$alcohol.bucket)
```

>  Above is the table summary for the alcohol bucket variable I created.

```{r}
ggplot(aes(x=quality), data = wine) + 
  geom_density(aes(color=alcohol.bucket)) +
    scale_color_brewer(palette='Greens',
    guide = guide_legend(title = 'Percent Alcohol', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +
    theme(panel.background = element_rect(fill='gray'), panel.grid.major = element_line(color='#969393'), panel.grid.minor = element_line(color='#969393'))
```

> From the density plot above, we can see that there are more high-alcohol wines with high quality ratings and more low-alcohol wines with the lower quality ratings.

```{r}
p1 <- ggplot(aes(x=quality, y=volatile.acidity), data = wine) + 
  geom_line(aes(color=alcohol.bucket), stat='summary', fun.y = median) +
    scale_color_brewer(palette='Greens',
    guide = guide_legend(title = 'Percent Alcohol', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +
    theme(panel.background = element_rect(fill='gray'), panel.grid.major = element_line(color='#969393'), panel.grid.minor = element_line(color='#969393'))

p2 <- ggplot(aes(x=volatile.acidity, y=quality), data = wine) + 
  geom_point(aes(color=alcohol.bucket)) +
    scale_color_brewer(palette='Greens', guide = guide_legend(title = 'Percent Alcohol', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +
    theme(plot.title = element_text(hjust = 0.5), panel.background = element_rect(fill='gray'), panel.grid.major = element_line(color='#969393'), panel.grid.minor = element_line(color='#969393'))

grid.arrange(p1,p2,ncol=1)
```

> Interestingly, wines in the highest alcohol bucket are the only ones that do not follow the trend of increasing quality with decreasing volatile acidity. Additionally, no wines in the two highest alcohol bucket had a quality rating of 3 and no wines in the lowest alcohol bucket had a quality rating of 8.

```{r}
# create buckets for volatile acidity
wine$volatile.acidity.bucket <- cut(wine$volatile.acidity, c(0,0.3,0.4,0.6,0.8,1.58))

table(wine$volatile.acidity.bucket)
```

>  Above is the table summary for the volatile acidity bucket variable I created.

```{r}
ggplot(aes(x=quality), data = wine) + 
  geom_density(aes(color=volatile.acidity.bucket)) +
    scale_color_brewer(palette='Purples',
    guide = guide_legend(title = 'Volatile Acidity', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +
    theme(panel.background = element_rect(fill='gray'), panel.grid.major = element_line(color='#969393'), panel.grid.minor = element_line(color='#969393'))
```

> Here, the density plot shows that there are more wines with low volatile acidity in the high quality bins and more wines with high volatie acidity in the low quality bins.

```{r}
p1 <- ggplot(aes(x=quality, y=alcohol), data = wine) + 
  geom_line(aes(color=volatile.acidity.bucket), stat='summary', fun.y = median) +
    scale_color_brewer(palette='Purples',
    guide = guide_legend(title = 'Volatile Acidity', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +
    theme(panel.background = element_rect(fill='gray'), panel.grid.major = element_line(color='#969393'), panel.grid.minor = element_line(color='#969393'))

p2 <- ggplot(aes(x=alcohol, y=quality), data = wine) + 
  geom_point(aes(color=volatile.acidity.bucket)) +
    scale_color_brewer(palette='Purples', guide = guide_legend(title = 'Volatile Acidity', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +
    theme(plot.title = element_text(hjust = 0.5), panel.background = element_rect(fill='gray'), panel.grid.major = element_line(color='#969393'), panel.grid.minor = element_line(color='#969393'))

grid.arrange(p1,p2,ncol=1)
```

> No wines in the lowest volatile acidity bucket had a quality rating of 3, but all volatile acidity buckets have some wines with a quality rating of 8. For the higher quality ratings, the higher volatile acidity buckets are associated with higher alcohol contents. Perhaps higher alcohol percentages help to compensate for any vinegar taste from high volatile acidity concentarations.


```{r}
ggplot(aes(x=quality.factor, y=volatile.acidity/alcohol), data=wine) +
  geom_boxplot()
```

```{r}
wine$volatile.acidity.alcohol.ratio <- wine$volatile.acidity/wine$alcohol

tapply(wine$volatile.acidity.alcohol.ratio, wine$quality.factor, summary)
```

> I wanted to see what the trend would be if I normalized volatile acidity with alcohol content. I expected this to be more level across the different quality ratings since it seems that high alcohol wines can still be high quality with high volatile acidity. However, the median of volatile acidity to alcohol decreases as quality increases (with similar values between 7 and 8), and the interquartile range decreases with increasing quality as well.

https://www.homebrewtalk.com/forum/threads/sulfate-to-chloride-ratio.526109/

```{r}
ggplot(aes(x=quality.factor, y=citric.acid/fixed.acidity), data=wine) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0,0.075))
```

```{r}
wine$citric.acid.fixed.acidity.ratio <- wine$citric.acid/wine$fixed.acidity

tapply(wine$citric.acid.fixed.acidity.ratio, wine$quality.factor, summary)
```

> Above, I plotted the ratio of citric acid to fixed acidity to see how it affects quality. This essentially normalizes the citric acid to the total amount of acids contributing to overall fixed acidity in the wine. We can see that the median ratio increases with increasing quality ratings, in a more linear fashion than the plot of citric acid alone. 

```{r}
ggplot(aes(x=quality.factor, y=sulphates/chlorides), data=wine) +
  geom_boxplot() +
  coord_cartesian(ylim=c(0,15))
```

```{r}
wine$sulphates.chloride.ratio <- wine$sulphates/wine$chlorides

tapply(wine$sulphates.chloride.ratio, wine$quality.factor, summary)
```

```{r}
cor(wine$quality, wine$sulphates.chloride.ratio, method='pearson')
```

> Above, I decided to look at how the sulphate to chloride ratio affects quality of wine. From research I found that some brewers believe the sulphate to chloride ratio is more important than the amounts themselves - this is generally discussed for beers but I decided to look at it for wines too. We can see that the sulphate to chloride ratio does appear to affect quality in a linear fashion - higher quality wines have higher median ratios.

```{r}
ggplot(aes(x=volatile.acidity, y=alcohol), data=wine) +
  geom_point(aes(color=quality.factor))+
    scale_color_brewer(palette='Blues',
    guide = guide_legend(title = 'Quality', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +
    theme(panel.background = element_rect(fill='gray'), panel.grid.major = element_line(color='#969393'), panel.grid.minor = element_line(color='#969393'))
```

> The plot above is another visualization showing that the higher quality wines are seen most at high alcohol content and low volatile acidity.

```{r}
ggplot(aes(x=volatile.acidity, y=citric.acid), data=wine) +
  geom_point(aes(color=quality.factor)) +
    scale_color_brewer(palette='Blues',
    guide = guide_legend(title = 'Quality', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +
    theme(panel.background = element_rect(fill='gray'), panel.grid.major = element_line(color='#969393'), panel.grid.minor = element_line(color='#969393'))
```

> From the plot above, we can see more high-quality wines at high citric acid contents and low volatile acidity. The linear relationship between citric acid and volatile acidity can also be seen again in this plot.

```{r}
ggplot(aes(x=volatile.acidity, y=citric.acid), data=wine) +
  geom_point(aes(color=quality.factor)) +
    scale_color_brewer(palette='Blues',
    guide = guide_legend(title = 'Quality', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +
    theme(panel.background = element_rect(fill='gray'), panel.grid.major = element_line(color='#969393'), panel.grid.minor = element_line(color='#969393')) +
  facet_wrap(~alcohol.bucket)
```

> When citric acid vs. volatile acidity is faceted over alcohol ranges, it appears that for the lower alcohol ranges, the higher qualities are seen at low volatile acidity and high citric acid concentration. For the higher alcohol ranges, the high quality wines are seen across a wider range of citric acid and volatile acidity values.

```{r}
m1 <- lm(quality ~ alcohol, data = wine)
m2 <- update(m1, ~ . + volatile.acidity)
m3 <- update(m2, ~ . + sulphates)
m4 <- update(m3, ~ . + chlorides)
m5 <- update(m4, ~ . + fixed.acidity)
m6 <- update(m5, ~ . + total.sulfur.dioxide)

mtable(m1, m2, m3, m4, m5, m6)
```

> Above, I've created a linear model to predict quality, using only the variables that made a difference to the R-squared value and that are not directly dependent on one another. Unfortunately, based on the R-squared value from m6, this linear model can only account for about 35% of the variance in wine quality.

```{r}
set.seed(42)
wineSample <- sample_n(wine, 1)
wineSample
```

```{r}
wineRandom <- data.frame(alcohol = 10.4, volatile.acidity = 0.64, fixed.acidity = 6.8, sulphates=0.58, chlorides=0.075, total.sulfur.dioxide=31)

predict(m6, newdata = wineRandom, interval="prediction", level = .95)

```

> For the above random wine sample with a quality of 6, my model predicted about 5.5, but 6 does fall within the 95% confidence interval. I want to try a high and low quality wine now to see how the model does.

```{r}
wineHQ <- subset(wine,wine$quality == 8)
head(wineHQ,1)
```

```{r}
wineHQ <- data.frame(alcohol = 12.8, volatile.acidity = 0.35, fixed.acidity = 7.9, sulphates=0.86, chlorides=0.078, total.sulfur.dioxide=37)

predict(m6, newdata = wineHQ,
                        interval="prediction", level = .95)
```

> The wine used here has a quality of 8 and my model predicted it to be about 6.7 based on its properties. 8 doesn't quite fall in the 95% confidence level but with rounding it would. 

```{r}
wineLQ <- subset(wine,wine$quality == 3)
head(wineLQ,1)
```

```{r}
wineLQ <- data.frame(alcohol = 9, volatile.acidity = 0.58, fixed.acidity = 11.6, sulphates=0.57, chlorides=0.074, total.sulfur.dioxide=47)

predict(m5, newdata = wineLQ, interval="prediction", level = .95)
```

> Above, with a low quality wine as a test (quality=3), we can see that the model did a worse job at predicting the quality. The lower end of the confidence interval is 3.9. Clearly, this is not a very effective model to predict wine quality, especially for wines that fall at the extremes of the quality spectrum. 

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \ investigation. Were there features that strengthened each other in terms of \ looking at your feature(s) of interest?

When plotting the ratio of citric acid to fixed acidity against quality, we can see a more linear relationship than with citric acid alone. This ratio, which normalizes the citric acid concentration to total fixed acidity, has a positive relationship with quality. 

Additionally, from plotting the ratio of sulphates to chlorides, I can see a more dramatic, more linear relationship with quality than the plot with sulphates alone. In fact, the ratio of these two variables has a higher correlation coefficient with quality than the two variables alone. This ratio is often said to be important for beer-brewing, so it's interesting that it appears to have an effect here as well. 

When placing the alcohol contents in buckets of ranges and plotting this along with other variables vs. quality, I saw that for most of the alcohol ranges, high quality was mainly associated with high citric acid and low volatile acidity. However, for the higher alcohol ranges, the high quality wines were seen across a wide range of citric acid and volatile acidity. This strengthens the conclusion that alcohol is the main contributor to quality, at least in this dataset. 

### Were there any interesting or surprising interactions between features?
I found it interesting that wines with high volatile acidity values can be high quality if they have high alcohol content. The wines in the higher alcohol ranges do not follow the trend for volatile acidity across quality levels like the wines in the lower alcohol ranges do. It seems that high alcohol percentages can compensate for the high volatile acidity levels. The trend seen between quality and the ratio of sulphates to chlorides as described above was also an interesting find.

### OPTIONAL: Did you create any models with your dataset? Discuss the \ strengths and limitations of your model.
I created a linear model with the dataset using 7 of the 11 features. Alcohol contributed the most to the model (R-squared value of 0.227), while the other features only increased it by small amounts. Based on the final R-squared value of 0.35 and the predictions from my test samples, it is clear that this is not a very effective model for predicting quality. For the more extreme quality ratings (3 and 8), it does not seem to do well at predicting the values even within a 95% confidence interval. Additionally, this model does not predict integer values. 
------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
p1 <- ggplot(aes(x=quality), data = wine) + 
  geom_density(aes(color=alcohol.bucket)) +
    scale_color_brewer(palette='Greens',
    guide = guide_legend(title = '% Alcohol Bucket', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +
    theme(plot.title = element_text(hjust = 0.5), panel.background = element_rect(fill='gray'), panel.grid.major = element_line(color='#969393'), panel.grid.minor = element_line(color='#969393')) +
  ggtitle('Density of Quality by Alcohol Content') +
  xlab('Quality') +
  ylab('Density')

p2 <- ggplot(aes(x=volatile.acidity, y=quality), data = wine) + 
  geom_point(aes(color=alcohol.bucket)) +
    scale_color_brewer(palette='Greens',
    guide = FALSE) +
    theme(plot.title = element_text(hjust = 0.5), panel.background = element_rect(fill='gray'), panel.grid.major = element_line(color='#969393'), panel.grid.minor = element_line(color='#969393')) +
  ggtitle('Quality by Volatile Acidity and Alcohol Content') +
  xlab('Volatile Acidity') +
  ylab('Quality')

grid.arrange(p1,p2,ncol=1)

```

### Description One
A greater proportion of wines from the highest alcohol range have higher qualities (7 or 8), and a greater proportion of wines from the lowest alcohol range have lower qualities (3 or 4) compared to the proportions from the distrubition of the other alcohol ranges. A switch in the order of proportions starts to begin between the quality ratings of 5 and 6. A much lower proportion of wines have a 3 or 8 rating than the other ratings overall. No wines were rated below 3 or above 8. In addition, higher quality ratings tend to be associated with lower volatile acidity, the exception being some wines in the highest alcohol range. 

### Plot Two
```{r echo=FALSE, Plot_Two}

p1 <- ggplot(aes(x=quality), data = wine) + 
  geom_density(aes(color=volatile.acidity.bucket)) +
    scale_color_brewer(palette='Purples',
    guide = guide_legend(title = 'Volatile Acidity Bucket (g/L)', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +
    theme(plot.title = element_text(hjust = 0.5), panel.background = element_rect(fill='gray'), panel.grid.major = element_line(color='#969393'), panel.grid.minor = element_line(color='#969393')) +
  ggtitle('Density of Quality by Volatile Acidity') +
  xlab('Quality') +
  ylab('Density')

p2 <- ggplot(aes(x=alcohol, y=quality), data = wine) + 
  geom_point(aes(color=volatile.acidity.bucket)) +
    scale_color_brewer(palette='Purples',
    guide = FALSE) +
    theme(plot.title = element_text(hjust = 0.5), panel.background = element_rect(fill='gray'), panel.grid.major = element_line(color='#969393'), panel.grid.minor = element_line(color='#969393')) +
  ggtitle('Quality by Alcohol Content and Volatile Acidity') +
  xlab('% Alcohol') +
  ylab('Quality')

grid.arrange(p1,p2,ncol=1)

```

### Description Two

A greater proportion of wines from the lowest volatile acidity range have higher qualities (7 or 8), and a greater proportion of wines from the highest volatile acidity range have lower qualities (3 or 4) compared to those from the other ranges. Just like with alcohol content, the switch in order of the proportions starts to occur between 5 and 6. The second plot confirms that high quality ratings tend to be associated with high alcohol content and low volatile acidity ranges, the exception being some wines with high alcohol content. 

### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(aes(x=quality.factor, y=sulphates.chloride.ratio), data=wine) +
  geom_boxplot(aes(fill=quality.factor)) + scale_fill_brewer(palette='Blues') +
  coord_cartesian(ylim=c(0,15)) +
  theme(plot.title = element_text(hjust = 0.5), legend.position='none', panel.background = element_rect(fill='gray'), panel.grid.major = element_line(color='#969393'), panel.grid.minor = element_line(color='#969393')) +
  ggtitle('Sulphate to Chloride Ratio by Quality Rating') +
  xlab('Quality') +
  ylab('Sulphate (g/L) to Chloride (g/L) Ratio')
```

### Description Three
Higher quality ratings are associated with higher sulphate to chloride ratios. It appears to be a somewhat linear, positive relationship.
------

# Reflection

This report explores data collected from 1599 red wine samples of the Portuguese Vinho Verde type around 2009. Eleven physiochemical properties were collected from each sample (input variables), along with a median quality rating from wine experts (target feature). I first looked at the distrubution of individual variables to get a better sense of th data. I then took a closer look at pairs of variables that appeared to be correlated, which eventually led me to creating multivariate plots to better explain quality. 

Percent alcohol content had the highest correlation with quality, with high quality wines generally being associated with relatively high alcohol percentages. There was also a clear trend between volatile acidity and quality; median volatile acidity decreased with increasing quality ratings. This was not surprising as volatile acidity is the property that can lead to an unpleasant vinegar taste in wine. Interestingly, it appears high alcohol content may be able to compensate for high volatile acidity in some cases. Citric acid and sulphate to chloride ratio are also correlated to quality, but this is a weaker correlation. Both have a positive relationship to quality. Surprisingly, sulfur dioxide content and residual sugar content do not seem to be correlated to quality, at least at the levels in this dataset. These were two features that I assumed would have a greater effect on taste, and therefore the quality rating. A relationship between two of the properties that I still do not understand is that of citric acid vs. volatile acidity. Volatile acidity (acetic acid content) decreases as citric acid content increases, and I was not able to come up with an explanation for this.

I constructed a linear model based on some of these correlations with quality, but my model only accounts for 35% of the variance in quality. My main struggle was coming up with other variables or transformation of variables to account for more of this variance. I think one of the main limitations is that this dataset mostly contains wines of mid-level quality ratings, i.e. there are very few wines with a low (3) or high(8) quality rating, and no wines below 3 or above 8. To investigate this data further, I would try to create a non-linear model to better predict quality, because the linear model I created was clearly ineffective for this particular dataset. A new dataset with a larger sample size with more wines of extreme quality would also help lead to a better predictive model. In addition, it would be interesting if tannins and/or carbon dioxide properties were added as features, since I would assume these would affect perceived quality. From what I read about Vinho Verde wines, many of these wines have a spritz to them from carbon dioxide content, but this was not mentioned in this dataset. These further explorations could lead to a better model for wine quality prediction. 

