Exploration of Red Wine Quality by Cady Patino
========================================================

> **Tip**: You will see quoted sections like this throughout the template to
help you construct your report. Make sure that you remove these notes before
you finish and submit your project!

> **Tip**: One of the requirements of this project is that your code follows
good formatting techniques, including limiting your lines to 80 characters or
less. If you're using RStudio, go into Preferences \> Code \> Display to set up
a margin line to help you keep track of this guideline!

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(gridExtra)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
wine <- read.csv('WineQualityReds.csv')
```

> This report explores a dataset containing 11 variables of the chemical properties of red wines and their corresponding quality rating. The quality rating for each wine is the median grade given by at least three wine experts and has a value between 0 (very bad) and 10 (very excellent).

# Univariate Plots Section

> **Tip**: In this section, you should perform some preliminary exploration of
your dataset. Run some summaries of the data and create univariate plots to
understand the structure of the individual variables in your dataset. Don't
forget to add a comment after each plot or closely-related group of plots!
There should be multiple code chunks and text sections; the first one below is
just to help you get started.

```{r echo=FALSE, Univariate_Plots}
str(wine)
```

```{r}
summary(wine)
```


> This data set contains 1599 observations of 13 variables. The first variable, X, appears to the index column. The next 11 variables are numbers pertaining to the chemical properties of wine, and the final variable is the integer rating for quality.

```{r}
qplot(x=quality, data=wine)
```

> From the above plot, it is clear that all quality values in this data set are integers. The majority of quality ratings in this data set are 5 or 6, and 3 and 8 are the least common ratings.


```{r}
qplot(x=fixed.acidity, data=wine, binwidth=0.1) +
  scale_x_continuous(breaks=seq(4,16,1))
```

```{r}
qplot(x=volatile.acidity, data=wine, binwidth=0.01) +
  scale_x_continuous(breaks=seq(0,1.6,0.1))
```

```{r}
qplot(x=citric.acid, data=wine, binwidth=0.01) +
  scale_x_continuous(breaks=seq(0,1,0.1))
```

> Above are the histograms related to acidity. The bulk of the fixed acidity data is between 6.5 and 9.5. The bulk of the volatile acidity is between 0.3 and 0.7. These two variables follow a fairly normal distribution with some outliers on the high end. The data for citric acid, on the other hand, is much more spread out, with peaks at 0 and about 0.49. The majority of citric acid data falls between 0 and 0.5.

```{r}
qplot(x=residual.sugar, data=wine, binwidth=0.1) +
  scale_x_continuous(breaks=seq(0,16,1))
```

```{r}
qplot(x=residual.sugar, data=wine, binwidth=0.01) +
  scale_x_log10()

```

> Above, I log transormed the long-tailed residual sugar data. On the log scale, since 3 is about halfway between 1 and 10, we can see that the bulk of the data falls around 2. The transformed data still has a tail due to outliers out to 16.

```{r}
qplot(x=chlorides, data=wine, binwidth=0.005) +
  scale_x_continuous(breaks=seq(0,0.62,0.05))
```

```{r}
qplot(log(x=chlorides), data=wine, binwidth=0.02)
```


```{r}
qplot(x=free.sulfur.dioxide, data=wine, binwidth=1) +
  scale_x_continuous(breaks=seq(0,72,5))
```

```{r}
qplot(x=log(free.sulfur.dioxide), data=wine, binwidth=0.05)
```


```{r}
qplot(x=total.sulfur.dioxide, data=wine, binwidth=2) +
  scale_x_continuous(breaks=seq(0,290,20))
```

```{r}
qplot(x=log(total.sulfur.dioxide), data=wine, binwidth=0.04)
```

```{r}
qplot(x=density, data=wine, binwidth=0.0001)
```

```{r}
qplot(x=pH, data=wine, binwidth=0.01) +
  scale_x_continuous(breaks=seq(0,4.01,0.1))
```

```{r}
qplot(x=sulphates, data=wine, binwidth=0.02) +
  scale_x_continuous(breaks=seq(0,2,0.2))
```

```{r}
qplot(x=log(sulphates), data=wine, binwidth= 0.02)
```


```{r}
qplot(x=alcohol, data=wine, binwidth= 0.1) +
  scale_x_continuous(breaks=seq(8,15,0.5))
```



# Univariate Analysis

> **Tip**: Now that you've completed your univariate explorations, it's time to
reflect on and summarize what you've found. Use the questions below to help you
gather your observations and add your own if you have other thoughts!

### What is the structure of your dataset?

### What is/are the main feature(s) of interest in your dataset?

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

### Did you create any new variables from existing variables in the dataset?

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?


# Bivariate Plots Section

> **Tip**: Based on what you saw in the univariate plots, what relationships
between variables might be interesting to look at in this section? Don't limit
yourself to relationships between a main output feature and one of the
supporting variables. Try to look at relationships between supporting variables
as well.

```{r echo=FALSE, Bivariate_Plots}

```

# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?


# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!